PanGraph: scalable bacterial pan-genome graph construction

### 2023-03-04 reviewer_response.tex

\response{The typos have been corrected.}

\reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}

\response{The capitalization has been corrected.}

\reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
% \Marco{here the pancontig size is the consensus sequence length. I would say that the fact that they have higher-than-average size is mostly due to synteny. Long blocks (several kbps) contain core genes that always appear in the same order, without any accessory sequence ever occurring in these regions. As opposed to accessory regions with a lot of structural diversity, that generates short fragments. But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks). Should we show block frequency distributions in the supplementary?}\\
% \Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size. Maybe pancontig `length' is clearer? I agree with you about synteny and think this is a nice point to add here to explain the observation.}
\response{The observation that core pancontigs are longer than average is in agreement with our prior the expectation since multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts. This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) pancontigs undergo.
      We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances.''}

@@

In the future we plan on providing a dedicated command to perform a more robust form of paralag splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph}.\\

We now make this explicit in SI sect.~{\SIalgo}D, that now contains these considerations.

% TODO: if this is sensible add a paragraph in SI about paralog splitting
% TODO: confirm with Nick the paralog split algorithm.
}
\Marco{Summarize, make more to the point. Say that this is now better discussed in SI.}

\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.". The value chosen for this parameter should be explicitly stated along with a rationale for the choice. It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
% \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
% \Marco{I agree. We set a minimal cutoff of 100bp. This is stated later in line 106. Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect. This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment. In the algorithm this minimal cutoff has different effects depending on the context. When aligning, if the match lenght is less than this threshold the merge is refused. For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block. The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as part of the merged block and not as separate small blocks. This might be explained in the SI?}\\
\response{The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph. This parameter controls the minimal size of pancontigs. In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment. For instance, an insertion could appear either as a separate pancontig, or as an insertion in the pancontig alignment, depending on its size. In practice this parameter is used in three ways:
      \begin{itemize}
            \item In addition to the pseudo-energy costraint, merges are refused if they have alignment length shorter than $\Lthr$.
            \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
            \item If a merger of two pancontigs would produce flanking overhangs shorter than $\Lthr$, these are not saved as separate pancontigs but are included in the alignment as flanking indels.
      \end{itemize}
      We choose a value of 100 bp in our analysis (and as standard value) to be able to detect variations at the gene scale, while encoding variations of up to few tens of bps in the alignments. This value can however be tuned by the user depending on the specific needs.\\
      We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph. To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo.\\
      % TODO: Homogenization in the main text with discussion on precision. Extend.
}

\reviewer{R3}{How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?}
% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address. Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI. The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks. Then consensus of the full alignment is re-evaluated. This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished. I don't think we ever quantified how much the consensus sequence degenerates during the iterations. Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer. To this end we added section {\SIalgo} in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
      In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable. However we do update the consensus of the pangontig at each merge using the majority rule on the merged alignment.\\
      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an eccessive amount of resources. However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly. This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
      % TODO: quantify this in simulations if there is time?
}

@@

% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
% TODO: core genome for simulated data if there is time?

%     "all core": 778,
%     "all softcore": 2081,
%     "all genes": 23128,
%     "all nondupl": 18798,
%     "drop core": 822,
%     "drop softcore": 2069,
%     "drop genes": 23054,
%     "drop nondupl": 18749

\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
\response{
      We appreciate the importance of the point raised by the reviewer.
      In it's current version \textit{PanGraph} does not support a post-processing refinement step in which alignments of adjacent edges of pancontigs are harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones.
      We agree with the reviewer that this post-processing smoothing of the graph would be an important improvement, and we consider adding it in subsequent versions of \textit{PanGraph}.
}

\reviewer{R3}{Overall the paper is well written. I like the basic methodology and the attention to using robust efficient existing algorithms and combining them to keep efficiency in mind. I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.}
% \Marco{if the fracturing derives from a real structural property of pangenomes (i.e. ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
\response{
      We thank the reviewer for their appreciation of the paper and of our methodology.\\
      We agree with the reviewer that fragmentation of the graph is an important concern, and that it can constitute a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph. For example datasets in which subsequent structural changes occurr at a scale comparable with $\Lthr$ cannot be correctly be transformed into a graph. Moreover an excessive number of pancontigs will also increase the amount of computational resources required.\\
      Within the size limit of our datasets, in the paper we investigated fragmentation in sect.~2B (``\textit{scaling with dataset size}''). In particular in fig.~5 we quantify how the size and number of pancontigs vary as a function of the number of isolates included in the graph. Here we observed good scaling properties, with pancontig size decreasing slowly as more isolates are added. We expect the threshold level of fragmentation to be reached around the point where the lines intercept the threshold length $\Lthr = 100$bp.\\

      % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure. We tested for excessive fragmentation in simulated data using partition entropy, comparing the real underlying structural partition to the one reconstructed by pangraph (not shown in the paper). We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel). Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in pancontigs having size close to $\Lthr$.\\
      % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
      % TODO: add this discussion in the text?
      % TODO: mention the partition entropy? This could be an argument against artificial fragmentation.
      % TODO: maybe add in the SI the block frequency distribution for different species/kernels?
}

\reviewer{R3}{An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene. I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs. How is this sustainable for species with large numbers of active IS elements? This could also be investigated for PanGrah using suitable simulated data.}
% \Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots). For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split. But I think this is a general problem and not something to investigate with simulations or fix here.}\\
% \Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
\response{
      We agree with the reviewer's observation that events such as the presence of an IS in an orthologous gene cluster would split the cluster in multiple pancontigs. For datasets in which isolates have a high amount of IS (longer than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
      This is in part an explicit design choice in \textit{PanGraph}. We plan to use this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be explcitly encoded in the graph. By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears, at the cost of fragmenting the graph. We are already using this property in a work in preparation \cite{shaw_preparation}, in which we compare the similarity between different contexts to obtain insights in the evolution of such sequences. This is especially important given the fact that IS often carry limited phylogenetic information, due to their limited size.\\
      We agree however with the reviewer in that this is not always a desirable feature. When not desired the user could be given the option to specify a list of IS sequences, that could be used to filter out mergers involving exclusively the IS. We consider adding this as a future improvement, however this requires further research and is not included in the current version of \textit{PanGraph}.
}

### 2023-03-04 reviewer_response.tex

\reviewer{R3}{This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.
      % \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained. Every SNP can generate a branching event. While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes. I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\
      One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
      \begin{itemize}
            \item \cite{angiuoli2011improving}
            \item \cite{angiuoli2011mugsy}
      \end{itemize}
      While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
      \begin{itemize}
            \item \cite{sutton2021pan}
            \item \cite{chan2015novel}
      \end{itemize}}
\responsedraft{}
\Marco{Copy form above and extend the answer}

\reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus. It captures two important issues: \textbf{aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum}. The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs. For initial alignments of single genomes to single genomes this doesn't matter for the most part because there is a single pancontig for each each genome and PGG context is mostly captured in the alignment. I am not so sure this is true as the PGG nodes fracture and become much smaller. When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence. I think this is a very important point as \textbf{distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it}. This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs. The structure is still useful even if this isn't done. Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
% \Marco{I agree, I think it's a different design choice. We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph. One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
\response{
We thank the reviewer for raising an important point that necessitates of more discussion in the paper. \textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
This is partially a design choice, as better explained in the answer to the last question. In short, the possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\

However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph. Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
The second method relies instead on structural information. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context.\\

### 2023-03-05 reviewer_response.tex

      The software was easy to install and the documentation available online is impressive.

      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\

      On reading I had a few queries, which could perhaps be clarified further in the text.}

\response{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}

@@

\reviewer{R2}{It is a bit depressing and unhelpful to structure a review only around problems and quibbles, so i wanted also to raise a number of
      Things I liked:

      \begin{enumerate}
            \item The paper is concise and well written, i really enjoyed reading it
            \item I liked the structure of the algorithms section, again i congratulate the authors, this is so much clearer than most papers. eg i was particularly happy to see the Guide Tree section start with your design constraints/aims.
            \item Fantastic choice of species in your empirical dataset - P. marinus diversity is insane.
            \item Line 215 - this is a remarkable achievement: ``Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added''
            \item Congratulations
      \end{enumerate}
}
\response{Thank you!}

\subsection*{Reviewer 3}

\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose. The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.\\
      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''. There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}

\response{
      We apologize if our previous discussion of the literature was incomplete, it was not our intention to downplay prior work.
      We thank the reviewer for highlighting additional important work and aspects that we previously didn't discuss.
      We have rewritten the part in the introduction to reflect these the work by these and other authors and added a few sentences to the discussion to further highlight previous and alternative approaches. We hope that these edits address the concerns of the reviewer.
}

@@

\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.}
% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
\response{We thank the reviewer for this important suggestion. We appreciate the need to include annotation in our pangenome graph structure. We created a script to decorate each annotation in a \texttt{gff} file with information on the pancontig location. The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. We plan on extending this in the future to use directly the pangenome graph as a reference and add annotations as features of pancontigs.
      % TODO: mention this in the text.
      % TODO: make the repository public.
}

\reviewer{R3}{
      This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
      One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
      \begin{itemize}
            \item \cite{angiuoli2011improving}
            \item \cite{angiuoli2011mugsy}
      \end{itemize}
      While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
      \begin{itemize}
            \item \cite{sutton2021pan}
            \item \cite{chan2015novel}
      \end{itemize}}
\response{
      We apologize and we modified the sentence in question. We also extendend the discussion of de Bruijn graphs methods in the introduction.\\
      We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
}
% \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained. Every SNP can generate a branching event. While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes. I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\

\reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus. It captures two important issues: aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum. The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs. For initial alignments of single genomes to single genomes this doesn't matter for the most part because there is a single pancontig for each each genome and PGG context is mostly captured in the alignment. I am not so sure this is true as the PGG nodes fracture and become much smaller. When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence. I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it. This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs. The structure is still useful even if this isn't done. Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
% \Marco{I agree, I think it's a different design choice. We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph. One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
\response{
We thank the reviewer for raising an important point that necessitates of more discussion in the paper. \textit{PanGraph} in its current state does not explicity distinguish between orthologs and paralogs.
This is partially a design choice, as better explained in the answer to the last question. In short, the possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\

@@

However, as the reviewer points out, in the case of distantly related paralogs, such a grouping might be undesirable as it introduces unneeded complications in the graph. Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
The second method relies instead on structural information. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context.\\

In the future we plan on providing a dedicated command to perform a more robust form of paralag splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig, to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph}.\\

We added SI sect.~{\SIalgo}D to include these considerations.

### 2023-03-05 main.tex
chore
update

\begin{abstract}
    The genomic diversity of microbes is commonly parameterized as single nucleotide polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
    However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
    Reference-based approaches are thus blind to the dynamics of the accessory genome, as well as variation within gene order and copy number.
    With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
    Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
    Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
    Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
\end{abstract}

@@

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}. Currently tools exist to build graphs containing thousands of isolates \cite{holley2020bifrost}, and have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}. However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome.

@@

Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        &          \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. We use the PanX pipeline to cluster homologous genes and characterize core genome size and diversity of our datasets. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$, corresponding to genes present in every isolate),
        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),
        and average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}

>>>>

\begin{abstract}
    The genomic diversity of microbes is commonly parameterized as single nucleotide polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
    However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
    Reference-based approaches are thus blind to the dynamics of the accessory genome, as well as variation within gene order and copy number.
    With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
    % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
    Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
    Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
    Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
\end{abstract}

@@

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}. Currently tools exist to build graphs containing thousands of isolates \cite{holley2020bifrost}, and have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}. However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome. Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.

@@

Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers. Alignment is performed on the base of sequence homology and does not rely on gene annotations.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss. Importantly, these two scales are naturally separated in our representation.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        &          \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Using the PanX analysis results we characterize the size and diversity of our dataset. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$, corresponding to genes present in every isolate),
        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),
        and average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}
